Generalized Bradley-Terry models

### 2025-04-22 inst/jss_hyper3_real.Rnw
typos

Further, one might follow~\cite{hankin2017_rnw_jss} and consider teams of
players.  In doubles tennis, for example, $\left\lbrace
p_1,p_2\right\rbrace$ might play $\left\lbrace p_3,p_4\right\rbrace$
and win with probability
$\frac{p_1+p_2}{p_1+p_2+p_3+p_4}$~\citep{hankin2010}.  It is natural
to combine this idea with that of a multiplicative factor.  For
example, suppose $p_1$ has just learned a new special technique, and
$p_4$ has just been given a performance-enhancing drug.  If so, we
might have likelihood function
%
\begin{equation}\label{like1234}
  \mathcal{L}{\left(p_1,p_2,p_3,p_4;\lambda,\mu\right)}\propto
  \frac{
    \left(\lambda p_1 + p_2\right)^{w_{12}}\cdot
    \left(p_3 + \mu p_4\right)^{w_{34}}
    }{
    \left(\lambda p_1+p_2+p_3+\mu p_4\right)^{w_{12} + w_{34}}
    }
\end{equation}
%
where $w_{12}$ and $w_{34}$ are the number of wins by each team
respectively.  Then $\lambda$ would represent the effectiveness of
$p_1$'s special technique, and $\mu$ the enhancement afforded to $p_4$
by the drug; values $>1$ would correspond to a positive effect.  In
this paper I generalize the \pkg{hyper2} R
package~\citep{hankin2017_rnw_jss}, available on CRAN~\citep{rcore2025},
for Bradley-Terry likelihoods to accommodate this class of
multiplicative factors with new \proglang{S3} \class{hyper3} and
associated methods.

@@

\subsection{The package in use}
Package idiom for creating a likelihood function for this dataset is
straightforward.  Considering first only Rangers-Celtic matches:
%
<<defineoldfirm, cache = TRUE>>=
oldfirm <- function(lambda){
    H <- hyper3(pnames = c("Rangers", "Celtic"))
    H[c(Rangers = lambda            )] %<>%      add(81     )
    H[c(                  Celtic = 1)] %<>%      add(     42)
    H[c(Rangers = lambda, Celtic = 1)] %<>% subtract(81 + 42)

    H[c(Celtic = lambda             )] %<>%      add(     76)
    H[c(                 Rangers = 1)] %<>%      add(46     )
    H[c(Celtic = lambda, Rangers = 1)] %<>% subtract(46 + 76)
    return(H)
}
@
%
Above we see package idiom for creating a log-likelihood function.  We
first create an empty {\tt hyper3} object with an appropriate names
attribute.  The next three lines, using \pkg{magrittr} assignment
pipes, give the results for Rangers at home, and the next three lines
give the result for Celtic at home.  It is then easy to create a
bespoke log-likelihood function for Rangers-Celtic matches:
%
<<useoldfirm, cache = TRUE>>=
oldfirm(1.88)
@
%
We see a log-likelihood function for all $82+42+76+46=245$
Rangers-Celtic matches that were not score draws, conditional on
assumed value $\lambda=1.88$.  This corresponds to the relevant terms
of Equation~\ref{firstfewRCFL}.  We may make inferences about the true
value of $\lambda$ using profile support:
%
<<profilelikeoldfirm, cache = TRUE>>=
f <- function(lambda){maxp(oldfirm(lambda), give = 1)$`Log-likelihood`}
lam <- seq(from = 0.9, to = 3, len = howmany)
logLike <- sapply(lam, f)
@
%
Thus \fct{f} returns the support at the evaluate, conditional on a
supplied value of $\lambda$.  We may use this to create a profile
likelihood (Figure~\ref{profilesupportlambda}), or to find the maximum
likelihood estimate for $\lambda$:

@@

The classic red bus-blue bus problem was originally stated by
\cite{mcfadden1980} although different metaphors have been employed
for the same phenomenon by \cite{tverskey1972_jss}.  The canonical problem
is as follows.  A commuter has a choice of going to work by different
transport methods.  Here, we can assume that he has possibly
strong---but unknown---views on different forms of transport which
might incorporate a random component.  If considering whether to use a
car or a bus, the colour of the bus is of no importance to him.  Such
preferences are problematic for Bradley-Terry: If our subject is
indifferent between cars and buses, then $p_{C} = p_{RB}$.  If he is
also indifferent to colour of bus, so $p_{RB}=p_{BB}$, the following
paradox occurs.  Bradley-Terry strengths must be equal so $p_{C} =
p_{RB} = p_{BB}=\frac{1}{3}$.  Thus $\Prob(C| \left\lbrace
C,RB\right\rbrace)=\frac{1}{2}$.  Now we introduce a blue bus as an
option.  Any realistic choice model will have $\Prob(C | \left\lbrace
C,RB,BB\right\rbrace)=\frac{1}{2}$ but Bradley-Terry gives us
$\frac{1}{3}$.  Various solutions to this problem have been proposed.
\cite{tverskey1972_jss}, for example, presents an approach involving
overlapping ``aspects'' of the elements of choice set;
\cite{mcfadden1980} suggests using a generalized extreme-value model.

Here I suggest an alternative in the form of a one-parameter
generalization of Plackett-Luce likelihood functions for ranking
options, using the multiplicative generalization to Bradley-Terry
strengths presented above.  As a synthetic example, we specify options
RB and BB for red bus and blue bus together options C, T and W for
car, train, and walking.

@@

We now apply the ideas above to a dataset obtained from the {\tt
  preflib} library~\citep{MaWa13a} in which ten different types of
sushi were placed in order of preference by 5000 judges.  Three of the
sushis (maguro, toro, tekka-maki) were made with tuna, and it is
reasonable to view these three types as being exchangeable in much the
same way as the red and blue buses.  To proceed, we need to define
equivalence classes of sushi---one for tuna-based sushis and one for
everything else.  Of course, one can also imagine the opposite
phenomenon whereby the judge decides he has had enough tuna for the
moment and this disincentivises further tuna choices.  This mechanism
would be modelled by having $\lambda<1$.  Implementation in
\code{hyper3} formalism is straightforward.  The dataset used is
\code{sushi\_table}, part of the \pkg{hyper2} package:
%
<<showsushitable>>=
noquote(head(sushi_table))
@
%
Thus, from the top line, we see that the first judge placed tamago
first, followed by anago and ika.  The equivalence classes are given
by
%
<<show_sushi_eq_classes>>=
sushi_eq_classes
@
%
so class 2 is the tuna group and class 1 everything else.  We can now
define a bespoke likelihood function for these observations, using
$\lambda$ as the argument:
%
<<defcalculate_sushi_H>>=
make_sushi_H <- function(lambda){
    H <- hyper3()
    for(i in seq_len(nrow(sushi_table))){
        H <- H + cheering3(sushi_table[i,], e = sushi_eq_classes, h = c(1, lambda))
    }
    return(H)
}
@
%
<<calclikesushi, echo = FALSE, cache = TRUE>>=
f <- function(lambda){
   H <- make_sushi_H(lambda)
   maxp(H, give = 1)$`Log-likelihood`
}
s <- exp(seq(from = log(0.9), to = log(10), len = howmany))
L <- sapply(s, f)
@ 
%
% chunk 'findstuffsushi' takes about 10-20 minutes to run
%
<<findstuffsushi, echo = FALSE, cache = TRUE>>=
osushi <- optimize(f, c(2, 4), maximum = TRUE)
@ 
%
% chunk tinysushi' takes about 5 minutes to run
%
<<tinysushi, echo = FALSE, cache = TRUE>>=
sushimaxlambda <- osushi$maximum
sushisupport <- osushi$objective - f(1)
@
%
% chunk 'findotherstuffsushi' takes about 10 minutes to run
%
<<findothersushistuff, echo = FALSE, cache = TRUE>>=
g <- function(lambda){f(lambda) - osushi$objective + 2}
sushi_lower <- uniroot(g, interval = exp(c(0, 1)), tol = 0.01)$root
sushi_upper <- uniroot(g, interval = exp(c(1, 2)), tol = 0.01)$root
@ 
%
(above, we note that the return value {\tt H} is the sum of successive
log-likelihood functions of each row of the table, assumed to be
independent).  Then, as before, we may define a profile likelihood
function of $\lambda$, Figure~\ref{profilesupportsushi}.  We
reject~$H_0\colon\lambda=1$ in favour of
$H_A\colon\lambda=\Sexpr{round(sushimaxlambda,2)}$ with about
$\Sexpr{round(sushisupport,2)}$ units of support.  Alternatively, we
may observe that this lies in the tail region of its asymptotic
$\chi^2_1$ distribution with a $p$~value of about
$\Sexpr{round(pchisq(sushisupport,df=1,lower.tail=FALSE),3)}$.  We can
therefore be reasonably sure that tuna sushi are somehow analagous to
red and blue buses, in the sense that all three tuna items tend to
occupy closely separated places in judges' preference ranks.

>>>>

Further, one might follow~\cite{hankin2017_rnw_jss} and consider teams of
players.  In doubles tennis, for example, $\left\lbrace
p_1,p_2\right\rbrace$ might play $\left\lbrace p_3,p_4\right\rbrace$
and win with probability
$\frac{p_1+p_2}{p_1+p_2+p_3+p_4}$~\citep{hankin2010}.  It is natural
to combine this idea with that of a multiplicative factor.  For
example, suppose $p_1$ has just learned a new special technique, and
$p_4$ has just been given a performance-enhancing drug.  If so, we
might have likelihood function
%
\begin{equation}\label{like1234}
  \mathcal{L}{\left(p_1,p_2,p_3,p_4;\lambda,\mu\right)}\propto
  \frac{
    \left(\lambda p_1 + p_2\right)^{w_{12}}\cdot
    \left(p_3 + \mu p_4\right)^{w_{34}}
    }{
    \left(\lambda p_1+p_2+p_3+\mu p_4\right)^{w_{12} + w_{34}}
    }
\end{equation}
%
where $w_{12}$ and $w_{34}$ are the number of wins by each team
respectively.  Then $\lambda$ would represent the effectiveness of
$p_1$'s special technique, and $\mu$ the enhancement afforded to $p_4$
by the drug; values $>1$ would correspond to a positive effect.  In
this paper I generalize the \pkg{hyper2} R
package~\citep{hankin2017_rnw_jss}, available on
CRAN~\citep{rcore2025}, for Bradley-Terry likelihoods to accommodate
this class of multiplicative factors with new \proglang{S3} class
\class{hyper3} and associated methods.

@@

\subsection{The package in use}
Creating a likelihood function for this dataset is straightforward.
As an illustration of package idiom, one way of representing a
likelihood function for Rangers-Celtic matches (here, ignoring draws)
would be:
%
<<defineoldfirm, cache = TRUE>>=
oldfirm <- function(lambda){
    H <- hyper3(pnames = c("Rangers", "Celtic"))
    H[c(Rangers = lambda            )] %<>%      add(81     )
    H[c(                  Celtic = 1)] %<>%      add(     42)
    H[c(Rangers = lambda, Celtic = 1)] %<>% subtract(81 + 42)

    H[c(Celtic = lambda             )] %<>%      add(     76)
    H[c(                 Rangers = 1)] %<>%      add(46     )
    H[c(Celtic = lambda, Rangers = 1)] %<>% subtract(46 + 76)
    return(H)
}
@
%
Above we see package idiom for creating a log-likelihood function.  We
first create an empty {\tt hyper3} object with an appropriate names
attribute.  The next three lines, using \pkg{magrittr} assignment
pipes, give the results for Rangers at home, and the next three lines
give the result for Celtic at home.  It is then easy to create a
bespoke log-likelihood function for Rangers-Celtic matches:
%
<<useoldfirm, cache = TRUE>>=
oldfirm(1.88)
@
%
We see a log-likelihood function for all $82+42+76+46=245$
Rangers-Celtic matches that were not score draws, conditional on
assumed value $\lambda=1.88$.  This corresponds to the relevant terms
of Equation~\ref{firstfewRCFL}.  We may make inferences about the true
value of $\lambda$ using profile support:
%
<<profilelikeoldfirm, cache = TRUE>>=
f <- function(lambda){maxp(oldfirm(lambda), give = 1)$`Log-likelihood`}
lam <- seq(from = 0.9, to = 3, len = howmany)
logLike <- sapply(lam, f)
@
%
Thus \fct{f} returns the support at the evaluate, conditional on a
supplied value of $\lambda$.  We may use this to create a profile
likelihood curve (Figure~\ref{profilesupportlambda}), or to find the
maximum likelihood estimate for $\lambda$:

@@

The classic red bus-blue bus problem was originally stated by
\cite{mcfadden1980} although different metaphors have been employed
for the same phenomenon by \cite{tverskey1972_jss}.  The canonical
problem is as follows.  A commuter has a choice of going to work by
different transport methods.  Here, we can assume that he has possibly
strong---but unknown---views on different forms of transport which
might incorporate a random component.  If considering whether to use a
car or a bus, the colour of the bus is of no importance to him.  Such
preferences are problematic for Bradley-Terry: If our subject is
indifferent between cars and buses, then $p_{C} = p_{RB}$.  If he is
also indifferent to colour of bus, so $p_{RB}=p_{BB}$, the following
paradox occurs.  Bradley-Terry strengths must be equal so $p_{C} =
p_{RB} = p_{BB}=\frac{1}{3}$.  Thus $\Prob(C| \left\lbrace
C,RB\right\rbrace)=\frac{1}{2}$.  Now we introduce a blue bus as an
option.  Any realistic choice model will have $\Prob(C | \left\lbrace
C,RB,BB\right\rbrace)=\frac{1}{2}$ but Bradley-Terry gives us
$\frac{1}{3}$.  Various solutions to this problem have been proposed.
\cite{tverskey1972_jss}, for example, presents an approach involving
overlapping ``aspects'' of the choice set; \cite{mcfadden1980}
suggests using a generalized extreme-value model.

Here I suggest an alternative in the form of a one-parameter
generalization of Plackett-Luce likelihood functions for ranking
options, using the multiplicative generalization to Bradley-Terry
strengths presented above.  As a synthetic example, we specify options
RB and BB for red bus and blue bus together with options C, T and W
for car, train, and walking.

@@

We now apply the ideas above to a dataset obtained from the {\tt
  preflib} library~\citep{MaWa13a} in which ten different types of
sushi were placed in order of preference by 5000 judges.  Three of the
sushis (maguro, toro, tekka-maki) were made with tuna, and it is
reasonable to view these three types as being exchangeable in much the
same way as the red and blue buses.  To proceed, we need to define
equivalence classes of sushi---one for tuna-based sushis and one for
everything else.  Of course, one can also imagine the opposite
phenomenon whereby the judge decides he has had enough tuna for the
moment and this disincentivises further tuna choices.  This mechanism
would be modelled by having $\lambda<1$.  Implementation in
\code{hyper3} formalism is straightforward.  The dataset used is
\code{sushi\_table}, part of the \pkg{hyper2} package:
%
<<showsushitable>>=
noquote(head(sushi_table))
@
%
Thus, from the top line, we see that the first judge placed tamago
first, followed by anago and ika.  The equivalence classes are given
by
%
<<show_sushi_eq_classes>>=
sushi_eq_classes
@
%
so class 2 is the tuna group and class 1 everything else.  We can now
define a bespoke likelihood function for these observations, using
$\lambda$ as the argument:
%
<<defcalculate_sushi_H>>=
make_sushi_H <- function(lambda){
    H <- hyper3()
    for(i in seq_len(nrow(sushi_table))){
        H <- H + cheering3(sushi_table[i,], e = sushi_eq_classes, h = c(1, lambda))
    }
    return(H)
}
@
%
<<calclikesushi, echo = FALSE, cache = TRUE>>=
f <- function(lambda){
   H <- make_sushi_H(lambda)
   maxp(H, give = 1)$`Log-likelihood`
}
s <- exp(seq(from = log(0.9), to = log(10), len = howmany))
L <- sapply(s, f)
@ 
%
% chunk 'findstuffsushi' takes about 10-20 minutes to run
%
<<findstuffsushi, echo = FALSE, cache = TRUE>>=
osushi <- optimize(f, c(1, 4), maximum = TRUE)
@ 
%
% chunk tinysushi' takes about 5 minutes to run
%
<<tinysushi, echo = FALSE, cache = TRUE>>=
sushimaxlambda <- osushi$maximum
sushisupport <- osushi$objective - f(1)
@
%
% chunk 'findotherstuffsushi' takes about 10 minutes to run
%
<<findothersushistuff, echo = FALSE, cache = TRUE>>=
g <- function(lambda){f(lambda) - osushi$objective + 2}
sushi_lower <- uniroot(g, interval = exp(c(0, 1)), tol = 0.01)$root
sushi_upper <- uniroot(g, interval = exp(c(1, 2)), tol = 0.01)$root
@ 
%
(above, we note that the return value {\tt H} is the sum of successive
log-likelihood functions of each row of the table, assumed to be
independent).  Then, as before, we may define a profile likelihood
function of $\lambda$, Figure~\ref{profilesupportsushi}.  We
reject~$H_0\colon\lambda=1$ in favour of
$H_A\colon\lambda=\Sexpr{round(sushimaxlambda,2)}$ with about
$\Sexpr{round(sushisupport,2)}$ units of support.  Alternatively, we
may observe that this lies in the tail region of its asymptotic
$\chi^2_1$ distribution with a $p$~value of about
$\Sexpr{round(pchisq(2*sushisupport,df=1,lower.tail=FALSE),3)}$.  We can
therefore be reasonably sure that tuna sushi are somehow analagous to
red and blue buses, in the sense that all three tuna items tend to
occupy closely separated places in judges' preference ranks.
